---
title: "Disease Incidence Forecasting in Kenya"
author: "Data Analyst"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Disease Incidence Forecasting Analysis

## 1. Load Required Packages
```{r}
# Core packages
library(tidyverse)
library(lubridate)
library(forecast)
library(ggplot2)
library(tseries)

# Additional packages for enhanced analysis
library(ggthemes)
library(plotly)
library(knitr)
library(kableExtra)
```

## 2. Data Loading and Preprocessing
```{r}
# Load the dataset
disease_data <- read.csv("../data/disease_incidence.csv")

# Data preprocessing
disease_data <- disease_data %>%
  mutate(
    date = as.Date(date),
    month = month(date, label = TRUE),
    year = year(date),
    quarter = quarter(date)
  )

# Summary statistics
summary_stats <- disease_data %>%
  group_by(region, disease_type) %>%
  summarise(
    mean_cases = mean(cases),
    median_cases = median(cases),
    sd_cases = sd(cases),
    min_cases = min(cases),
    max_cases = max(cases)
  )

# Display summary statistics
kable(summary_stats, caption = "Summary Statistics by Region and Disease Type") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 3. Exploratory Data Analysis
```{r}
# Time series plot with trend line
ggplot(disease_data, aes(x = date, y = cases)) +
  geom_line(color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  labs(title = "Disease Incidence Over Time with Trend",
       x = "Date",
       y = "Number of Cases") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Seasonal plot
ggplot(disease_data, aes(x = month, y = cases, group = year)) +
  geom_line(aes(color = factor(year))) +
  labs(title = "Seasonal Pattern of Disease Incidence",
       x = "Month",
       y = "Number of Cases",
       color = "Year") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Box plot by region
ggplot(disease_data, aes(x = region, y = cases, fill = region)) +
  geom_boxplot() +
  labs(title = "Distribution of Cases by Region",
       x = "Region",
       y = "Number of Cases") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## 4. Time Series Analysis
```{r}
# Create time series object
ts_data <- ts(disease_data$cases, frequency = 12)

# Seasonal decomposition
decomp <- stl(ts_data, s.window = "periodic")
autoplot(decomp) +
  labs(title = "Seasonal Decomposition of Time Series") +
  theme_minimal()

# ACF and PACF plots
par(mfrow = c(2,1))
acf(ts_data, main = "Autocorrelation Function")
pacf(ts_data, main = "Partial Autocorrelation Function")
```

## 5. Model Building
```{r}
# Split data into training and testing sets
train_size <- floor(0.8 * length(ts_data))
train_data <- ts_data[1:train_size]
test_data <- ts_data[(train_size + 1):length(ts_data)]

# Fit ARIMA model
arima_model <- auto.arima(train_data)

# Model summary
summary(arima_model)

# Check model assumptions
checkresiduals(arima_model)
```

## 6. Model Evaluation
```{r}
# Make predictions
forecast_values <- forecast(arima_model, h = length(test_data))

# Calculate accuracy metrics
accuracy_metrics <- accuracy(forecast_values, test_data)
kable(accuracy_metrics, caption = "Model Accuracy Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Plot forecast
autoplot(forecast_values) +
  autolayer(test_data, series = "Actual") +
  labs(title = "Disease Incidence Forecast",
       x = "Time",
       y = "Number of Cases") +
  theme_minimal()
```

## 7. Future Forecast
```{r}
# Generate 12-month forecast
future_forecast <- forecast(arima_model, h = 12)

# Plot future forecast
autoplot(future_forecast) +
  labs(title = "12-Month Disease Incidence Forecast",
       x = "Time",
       y = "Number of Cases") +
  theme_minimal()

# Print forecast values
kable(future_forecast$mean, caption = "12-Month Forecast Values") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 8. Conclusion
The analysis reveals several key findings:

1. **Trend Analysis**: The data shows a `r if(mean(diff(ts_data)) > 0) "positive" else "negative"` trend in disease incidence over time.

2. **Seasonal Patterns**: There are clear seasonal patterns in the data, with `r names(which.max(table(disease_data$month)))` typically showing the highest incidence.

3. **Model Performance**: The ARIMA model demonstrates `r if(accuracy_metrics[2] < 0.1) "good" else "moderate"` forecasting accuracy, with a MAPE of `r round(accuracy_metrics[2] * 100, 2)`%.

4. **Forecast Implications**: The 12-month forecast suggests `r if(mean(diff(future_forecast$mean)) > 0) "an increasing" else "a decreasing"` trend in disease incidence.

These findings can help public health officials in Kenya to:
- Plan resource allocation
- Implement timely interventions
- Monitor disease trends
- Prepare for potential outbreaks 